---
title: "EDA"
author: "Ruoying Deng"
date: "2024-12-03"
output: html_document
---
```{r}
#necessary library
library(tibble)
library(kableExtra)
library(ggplot2)
library(dplyr)
library(ggcorrplot)
library(patchwork)
```

```{r}
#clean clinical attributes
cl_clinical_dat <- clinical_dat %>%
  select(-"cellularity",-"tumor_stage",-"X3.gene_classifier_subtype")
```



```{r}
variables <- c("age_at_diagnosis", "lymph_nodes_examined_positive", 
               "mutation_count", "nottingham_prognostic_index", 
               "overall_survival_months", "tumor_size")

# Create individual density plots for each variable grouped by 'overall_survival'
plots <- lapply(variables, function(var) {
  ggplot(data = clinical_dat, aes_string(x = var, fill = "overall_survival")) +
    geom_density(alpha = 0.5, color = "black") +
    scale_fill_manual(values = c("0" = "red", "1" = "blue")) +
    theme_minimal() +
    labs(
      title = paste("Density Plot of", var, "by overall_survival"),
      x = var,
      y = "Density",
      fill = "overall_survival"
    )
})


combined_plot <- wrap_plots(plots, ncol = 2)
print(combined_plot)

```
For the distribution of all numerical data, some of them are normally distributed like age_at_diagnosis, but most of the features are right skewed with a lot of outliers like lymph_nodes_examined_positive, mutation_count, and tumor_size. therefore should be transformed. Logarithm transformations were used to modify the data. The final data set contains 1604 samples with 6 numeric and 24 categorical variables.

```{r}
cl_clinical_dat$mutation_count_log <- log1p(cl_clinical_dat$mutation_count) # log1p handles zeros
cl_clinical_dat$lymph_nodes_examined_positive_log <- log1p(cl_clinical_dat$lymph_nodes_examined_positive)
cl_clinical_dat$tumor_size_sqrt <- sqrt(cl_clinical_dat$tumor_size) #moderate skewed

# Prepare data in long format
transformed_data <- cl_clinical_dat %>%
  select(mutation_count_log, lymph_nodes_examined_positive_log, tumor_size_sqrt) %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Value")

# Plot density for transformed variables
ggplot(transformed_data, aes(x = Value, fill = Variable)) +
  geom_density(alpha = 0.5) +
  theme_minimal() +
  labs(
    title = "Density Plots for Transformed Variables",
    x = "Transformed Values",
    y = "Density"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.title = element_text(size = 12),
    legend.position = "top"
  )

```

```{r}
clinical_data_scaled <- cl_clinical_dat %>%
  select_if(is.numeric) %>% 
  scale() %>% 
  as.data.frame()
#remove patient_id
clinical_data_scaled<-clinical_data_scaled%>%
  select(-"patient_id")
# Reshape the data
clinical_data_melted <- clinical_data_scaled %>%
  rownames_to_column("Observation") %>%
  pivot_longer(-Observation, names_to = "Attribute", values_to = "Value")

# Create the horizontal boxplot
ggplot(clinical_data_melted, aes(x = Value, y = Attribute)) +
  geom_boxplot(aes(fill = Attribute), outlier.color = "black", outlier.shape = 16) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 0, vjust = 1, hjust = 1),
    axis.title.y = element_text(angle = 0)
  ) +
  labs(
    title = "The Distribution of Numeric Clinical Attributes in the Dataframe",
    x = "Range after Standardization",
    y = "Attribute"
  )
```
As we can see, after transformation for mutation_count, lymph_nodes_examined_positive, and tumor size. The long tail in the original variable is minimized, making the distribution more balanced.

```{r}
# Identify categorical variables
categorical_dat <- clinical_dat %>%
  select_if(is.character) %>%    
  bind_cols(clinical_dat %>% select_if(is.factor))

# Frequency and proportion by survival
categorical_summary <- categorical_dat %>%
  pivot_longer(cols = -overall_survival, names_to = "Variable", values_to = "Category") %>%
  group_by(Variable, Category, overall_survival) %>%
  summarise(Count = n(), .groups = "drop") %>%
  mutate(Proportion = Count / sum(Count))

# Print summary
print(categorical_summary)

```

```{r}
# Loop through categorical variables for stacked bar plots
categorical_vars <- c("cancer_type", "type_of_breast_surgery", "cancer_type_detailed","pr_status")

plots <- lapply(categorical_vars, function(var) {
  ggplot(clinical_dat, aes(x = .data[[var]], fill = as.factor(overall_survival))) +
    geom_bar(position = "fill", alpha = 0.8) +
    scale_fill_manual(values = c("0" = "red", "1" = "skyblue"), labels = c("Died", "Survived")) +
    theme_minimal() +
    labs(
      title = paste("Distribution of", var, "by Overall Survival"),
      x = var,
      y = "Proportion",
      fill = "Overall Survival"
    )
})

# Display the plots
wrap_plots(plots,ncol = 2)

```


```{r}
# Calculate the correlation matrix 
correlation_matrix <- cor(clinical_data_scaled, use = "complete.obs")

# Create the correlation matrix plot
ggcorrplot(correlation_matrix, method = "circle", 
           lab = TRUE, lab_size = 3, 
           title = "Correlation Matrix", 
           colors = c("red", "white", "blue"), 
           ggtheme = theme_minimal())
  labs(title = "Correlation Heatmap", x = "Variables", y = "Variables")
```



```{r}
ggplot(data = clinical_dat, aes(x = tumor_size, y = tumor_stage, fill = as.factor(overall_survival))) +
  geom_boxplot(alpha = 0.7, outlier.color = "black", outlier.shape = 16) +
  scale_fill_manual(values = c("0" = "salmon", "1" = "skyblue"), name = "Overall Survival") +
  theme_minimal() +
  labs(
    title = "Distribution of Tumor Size by Tumor Stage and Survival",
    x = "Tumor Size",
    y = "Tumor Stage"
  ) +
  theme(
    legend.position = "right",
    axis.text.y = element_text(size = 12),
    axis.text.x = element_text(size = 12)
  )
```
As the Tumer stage increases the tumor size increases as well. Also, if lower tumor stages the probability of survival is higher than when the patient reaches the fourth stage

```{r}
treatment_vars <- c("chemotherapy", "hormone_therapy", "radio_therapy")

# Create bar plots for each treatment variable grouped by survival
plots <- lapply(treatment_vars, function(var) {
  ggplot(data = categorical_dat, aes_string(x = var, fill = "as.factor(overall_survival)")) +
    geom_bar(position = "stack", alpha = 0.8) +
    scale_fill_manual(values = c("0" = "red", "1" = "skyblue"), labels = c("0" = "Died","1"="Survived") , name = "Survival Status") +
    theme_minimal() +
    labs(
      title = NULL,
      x = var,
      y = "Number of Patients"
    )
})



combined_plot <- wrap_plots(plots, ncol = 3) +
  plot_annotation(
    title = "The Distribution of Treatment and Survival",
    theme = theme(plot.title = element_text(hjust = 0.5, size = 14))
  )

# Display the combined plot
print(combined_plot)
```
Hormonal therapy appears to be associated with higher survival rates, as most patients who received it survived.
The group of patients who did not receive chemotherapy is larger and has a better survival rate, but this might reflect patient selection or other confounding factors.
```{r}
# Summarize survival rates by each treatment type
survival_summary <- categorical_dat %>%
  pivot_longer(cols = c("chemotherapy", "hormone_therapy", "radio_therapy"),
               names_to = "Treatment_Type",
               values_to = "Treatment_Status") %>%
  group_by(Treatment_Type, Treatment_Status, overall_survival) %>%
  summarise(Count = n(), .groups = "drop") %>%
  mutate(Percentage = Count / sum(Count) * 100)%>%

print(survival_summary)

```
```{r}
# Create a contingency table for survival and each treatment type
chi_test_results <- lapply(c("chemotherapy", "hormone_therapy", "radio_therapy"), function(var) {
  treatment_table <- table(categorical_dat[[var]], categorical_dat$overall_survival)
  chi_test <- chisq.test(treatment_table)
  list(Treatment = var, P_Value = chi_test$p.value)
})

# Convert results to a data frame
chi_test_results_df <- do.call(rbind, chi_test_results)
print(chi_test_results_df)

```
The p-value for radio_therapy is well below 0.05,indicates a strong and statistically significant association between radio therapy and survival.
