---
title: "Data Preprocessing"
output: html_document
date: "2024-12-03"
---
```{r}
library(tidyverse)
library(tibble)
library(kableExtra)
```

```{r}
dat <- read.csv("/Users/viv/Desktop/Survival Analysis/METABRIC_RNA_Mutation.csv")
```


```{r}
dat[dat == ""] <- NA  
dat_cleaned <- dat %>% 
  mutate(type_of_breast_surgery = as.factor(na_if(type_of_breast_surgery, "")),
         cancer_type = as.factor(cancer_type), 
         cancer_type_detailed = as.factor(na_if(cancer_type_detailed, "")),
         cellularity = factor(na_if(cellularity, ""), levels = c("Low", "Medium", "High")),
         chemotherapy = as.factor(chemotherapy),
         pam50_._claudin.low_subtype = as.factor(pam50_._claudin.low_subtype),
         cohort = as.factor(cohort),
         er_status_measured_by_ihc = as.factor(na_if(er_status_measured_by_ihc, "")),
         er_status = as.factor(er_status), 
         neoplasm_histologic_grade = as.factor(neoplasm_histologic_grade),
         her2_status_measured_by_snp6 = factor(na_if(her2_status_measured_by_snp6, "UNDEF"), levels = c("LOSS", "NEUTRAL", "GAIN")),
         her2_status = as.factor(her2_status),
         tumor_other_histologic_subtype = as.factor(na_if(tumor_other_histologic_subtype, "")),
         hormone_therapy = as.factor(hormone_therapy),
         inferred_menopausal_state = factor(na_if(inferred_menopausal_state, ""), levels = c("Pre", "Post")),
         integrative_cluster = as.factor(integrative_cluster),
         primary_tumor_laterality = as.factor(na_if(primary_tumor_laterality, "")),
         oncotree_code = as.factor(na_if(oncotree_code, "")),
         overall_survival = as.factor(overall_survival),
         pr_status = as.factor(pr_status),
         radio_therapy = as.factor(radio_therapy),
         X3.gene_classifier_subtype = as.factor(X3.gene_classifier_subtype),
         tumor_stage = as.factor(tumor_stage),
         death_from_cancer = as.factor(na_if(death_from_cancer, "")))

write.csv(dat_cleaned, "dat_cleaned.csv", row.names = FALSE)
```

the dataset includes 31 clinical attributes, m-RNA levels z-score for 331 genes, and mutation in 175 genes for 1904 breast cancer patients

```{r}
#clinical attributes
clinical_dat <-select(dat_cleaned,(1:31))
summary(clinical_dat)
```

```{r}
clinical_attributes <- tribble(
  ~Category,                ~Attributes,
  "General Info",           "patient_id, age_at_diagnosis, inferred_menopausal_state, primary_tumor_laterality",
  "Patient Status",         "overall_survival, overall_survival_months, death_from_cancer",
  "Tumor Info",             "pam50_._claudin.low_subtype, cellularity, er_status, neoplasm_histologic_grade, her2_status, pr_status, X3_gene_classifier_subtype, tumor_size",
  "Cancer Type",     "cancer_type, cancer_type_detailed, tumor_other_histologic_subtype, integrative_cluster, oncotre_code",
  "Treatment",              "type_of_breast_surgery, chemotherapy, hormone_therapy, radio_therapy"
)

kable(clinical_attributes, col.names = c("Category", "Attributes")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F) %>%
  column_spec(2, width = "30em")
```

```{r}
# Calculate missing values and percentages
missing_summary <- clinical_dat %>%
  summarise_all(~ sum(is.na(.))) %>%
  t() %>%
  as.data.frame()

missing_summary$Percentage <- (missing_summary$V1 / nrow(clinical_dat)) * 100
colnames(missing_summary) <- c("Missing_Values", "Missing_Percentage")

missing_summary <- missing_summary %>%
  rownames_to_column(var = "Feature") %>%
  arrange(desc(Missing_Percentage))
          
view(missing_summary)

# Plot missing data percentages
ggplot(missing_summary, aes(x = reorder(Feature, -Missing_Percentage), y = Missing_Percentage)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +
  labs(
    title = "Percentage of Missing Data by Feature",
    x = "Feature",
    y = "Missing Percentage (%)"
  ) +
  theme_minimal()
```
The missing values were defined with an empty cell in this data set. After replacing the empty cell with NA in the data set distribution of missing values for each variable in clinical dataset was presented here: 
assuming less than five percent missing values are an acceptable threshold,The cellularity has missing percentage of 40%, 26% missing percent in tumor_stage and 11% missing percent in X3.gene_classifier_subtype. These three variables present redundant data and could be removed.
```{r}
mrna_z_scores <- dat_cleaned %>%
  select_if(is.numeric)%>%
  select(-c(2:14))
view(mrna_z_scores)

missing_summary2 <- mrna_z_scores%>%
 summarise_all(~ sum(is.na(.)))%>%
  t() %>%
  as.data.frame()

missing_summary2 <- missing_summary2 %>%
  tibble::rownames_to_column(var = "Feature") %>%
  rename(Missing_Values = V1) %>%
  arrange(desc(Missing_Values))

# Extract Mutation Data for 175 Genes
mutation_data <- dat_cleaned %>%
  select(matches("mut"))
```

The dataset for m-RNA levels z-score and mutation in dna is complete.

```{r}
# # Clinical vars
# colnames(dat[,1:31])
# #colnames(dat[,32:ncol(dat)])
# 
# subset_columns <- dat[, 32:ncol(dat)]
# 
# # Function to calculate variance and check against threshold
# near_zero_variance <- function(column, threshold = 0.0001) {
#   if (is.numeric(column)) { # Check if the column is numeric
#     return(var(column) < threshold) # Check variance
#   } else {
#     return(FALSE) # Non-numeric columns are skipped
#   }
# }
# 
# # Apply the function to each column
# nzv_flags <- sapply(subset_columns, near_zero_variance)
# 
# # Identify columns with zero or near-zero variance
# nzv_columns <- names(nzv_flags[nzv_flags])
# 
# # no zero variance or near zero variance columns
# print(nzv_columns)
# 
# # no missing data in any column
# sum(is.na(dat[, 32:ncol(dat)]))





